---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Set global options.
```{r}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.height=6, fig.width=6)

```

Download the file using the specified link and unzip in the working directory.  

```{r eval=FALSE}
setwd("C:/Users/ssbhat3/Desktop/Coursera ReproducibleResearch/RepData_PeerAssessment1/RepData_PeerAssessment1")
getwd()
Source <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
Destination <- "activity.zip"
download.file(Source, Destination, mode="wb") 
unzip(Destination)
```

This creates a file `activity.csv`. Read into a `data.table`.

```{r}
library(data.table)
DTact <- as.data.table(read.csv("activity.csv"))
DTact
```

Set the date in a format that `R` recognizes.

```{r results='hide'}
DTact[, lapply(.SD, class)]
DTact[, date := as.Date(date)]
str(DTact)
```

We are now all set to process data.

## What is mean total number of steps taken per day?

Calculate the number of steps per day.
```{r results='hide'}
DTact[, Total := sum(steps), by=date]

```

The histogram reveals the central tendency and spread.

```{r}
library(ggplot2)
DTact[, {
  m <- ggplot(.SD, aes(x=Total))
  m <- m + geom_histogram(aes(y=..density..), binwidth=3000, 
                          color="black", fill="wheat", alpha=0.2) 
  m <- m + geom_density(fill="gray", alpha=0.2)
  m <- m + geom_vline(aes(xintercept=mean(Total, na.rm=TRUE)),
                      color="red", linetype="dashed", size=1)
  m <- m + geom_vline(aes(xintercept=median(Total, na.rm=TRUE)),
                      color="blue", linetype="dashed", size=1)
  m <- m + xlab("Daily Activity (Steps)") + ylab("Density") + 
          ggtitle("Variation in Daily Activity Levels")
  
}]
```

The mean and median are shown on the histogram.

```{r results='hide'}
DTact[, summary(Total, na.rm=TRUE)]
```

The total number of steps per day have `mean=10770` and `median=10760`. The overlay shows data are very noisy.

## What is the average daily activity pattern?

Let's look at any pattern in activity by time of day. The time of day is specified by interval number. We can average the number of steps in any interval across all days.

```{r}
DTint <- DTact[, .(interMean = mean(steps, na.rm=TRUE)), by=interval]
DTint[, {
  t <- ggplot(.SD, aes(interval, interMean))
  t <- t + geom_line()
  t <- t + geom_smooth()
  t <- t + xlab("Interval") + ylab("Avg. Activity Level (Steps)") + 
          ggtitle("Daily Activity Pattern (Average)")
}]
```

The time-series reaffirm noisy data. 

```{r results='hide'}
DTint[, Smoothed := {
  myLoess <- loess(data=.SD, interMean ~ interval)
  Smoothed <- myLoess$fitted
}]
DTint[, interval[c(which.max(Smoothed), which.max(interMean))]]
```

The maximum number of steps across all days is found in interval `835` in raw data. Given the data are noisy, smoothing may produce a more robust estimate. Applying loess method, the maxima is found in interval `900`.

## Imputing missing values

There are a number of records with missing values. 

```{r}
DTact[, sum(is.na(steps))]
```

There are `2304` records missing steps. Replace these with the median value in that interval. For this purpose, I have written a function `impute`. Then tweaked it for vectorization with `Vectorize`.

```{r}
impute <- function(x, y) {
  if (is.na(x)) {
    x <- y
  }
  x
}
vImpute <- Vectorize(impute)
```

The function replaces `NA` values with the median value in that interval.

```{r results='hide'}
DTact[, interMedian := median(steps, na.rm=TRUE), by=interval]
DTact[, steps := vImpute(steps, interMedian)]
DTact[, interMedian := NULL]
```

Recompte totals with imputed values. Redo the histogram.

```{r}
DTact[, Total := sum(steps), by=date]
DTact[, {
  m <- ggplot(.SD, aes(x=Total))
  m <- m + geom_histogram(aes(y=..density..), binwidth=3000, 
                          color="black", fill="wheat", alpha=0.2) 
  m <- m + geom_density(fill="gray", alpha=0.2)
  m <- m + geom_vline(aes(xintercept=mean(Total, na.rm=TRUE)),
                      color="red", linetype="dashed", size=1)
  m <- m + geom_vline(aes(xintercept=median(Total, na.rm=TRUE)),
                      color="blue", linetype="dashed", size=1)
  m <- m + xlab("Daily Activity (Steps)") + ylab("Density") + 
          ggtitle("Variation in Daily Activity Levels")
  
}]

DTact[, summary(Total)]
```

Imputation of values yields a mean of `9504` and median of `10400` steps per day. These adjusted values are lower than before.

## Are there differences in activity patterns between weekdays and weekends?

The function `wDay` determines whether a specified date is a weekday or a weekend. It's vectorized version is `vwDay`.

```{r}
wDay <- function(date) {
  day <- weekdays(date)
  if (day %in% c("Saturday", "Sunday") ) {
    return("weekend")
  } else {
    return("weekday")
  }
}
wDay(as.Date("2015-10-18"))
vwDay <- Vectorize(wDay)
```

Compute a new factor variable to indicate weekend. The time-series plot shows the activity on weekdays and weekends. 

```{r}
DTact[, day := factor(vwDay(date))]
DTwee <- DTact[, .(interMean = mean(steps, na.rm=TRUE)), 
      by=c("interval", "day")]
DTwee[, {
  t <- ggplot(.SD, aes(interval, interMean))
  t <- t + geom_line()
  t <- t + geom_smooth()
  t <- t + facet_grid(day ~ .)
  t <- t + ylab("Activity Level (Steps)") + 
          ggtitle("Activity Levels: Weekday v Weekend")
}]
```

The plot shows difference in daily activity between weekday and weekend. We can fit a regression model to examine the assocation between day of the week and activity level. A regression model can show statistically significant differences from day to day.

```{r}
wDay <- DTact[, {
  lm(Total ~ day)         
}]
summary(wDay)
```

This  model shows that daily activity on a weekend is higher than the activity on a weekday and the difference is statistically significant at 95% confidence. The boxplot illustrates this result. 

```{r}
DTact[, lapply(.SD, class)]
DTact[, {
  p <- ggplot(.SD, aes(x=day, y=Total, fill=factor(day)))
  p <- p + geom_boxplot() + geom_smooth(method="lm")
  p <- p + ylab("Activity Level (Steps)") + 
          ggtitle("Activity Level: Weekday v Weekend")
}] 
```

## Conclusion

Data are noisy with over 10% records missing values. The imputation approach produces a preponderance of zero values. A more sophisticated approach may need to be considered. 

Regression shows that weekend affects the daily activity level and that the difference is statistically significant. It may be worthwhile looking closer at this association. A regression model with all 7 days of the week as the explanatory variables should be illuminatiing. 